---
title: "Analysis of NOAA Severe Weather Event Data"
author: "Sintecs"
date: "August 20, 2015"
output: html_document
---
#An Analysis of Injury and Economic Impact of Weather

##Synopsis
This report is an examination and analysis of the NOAA data from 1950 to November 2011.  The recordset used contains 902297 observations of 37 variables.  The variables chosen in this analysis for examination are EVTYPE, FATALITIES, INJURIES, CROPDMGEXP, PROPDMGEXP, CROPDMG and PROPDMG to determine the impact that each event had and determine which were the two worst impacting event types.  The worst physical human impact is from tornadoes with excessive heat being second worst in death and thunderstorm winds in injury impact.  The worst events for property damaage are floods and hurricanes/typhoons.

##Data Processing
The raw data is loaded from [here][2] and the bz2 file is decompressed and loaded into a variable called weather.data in R.  From here we cleanup the data in the EVTYPE column to standardize misspellings and errors in notation of the data.  We then examine the data in a series of plots to show the events with most human impact (injuries and deaths) and economic impact as a total of economic and crop impact severity.
```{r load the dataset}
knitr::opts_chunk$set(cache = TRUE)
## URL of the dataset and filename
sourceDataURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

## Temporary File for storing the downloaded zip and extract the files we want into R
destfilename <- "repdata_data_StormData.csv.bz2"
if (!file.exists(destfilename))
{
    download.file(sourceDataURL, destfilename, method = "libcurl")
}
weather.data <- read.csv(destfilename)
```
##Load and aggregate the downloaded data
###Cleanup the EVTYPE column using grep
Though the current [Storm Data Documentation][1] book in section 2.1.1, Storm Data Event Table denotes only 48 different event types, the raw data contains, as of this writing, 985 events.   

As part of pre-processing of this file we will use grep and gsub with regular expressions to replace errors in the names (spelling, punctuation and etc.) with the correct event name from the Storm Data Event Table.  Also, items where there are characters after a known event name will have those characters stripped so the Storm Data Event Table name is used.  If there is an event where the name contains two (or more) values from the Storm Data Event Table the first one noted will be used.

First we will find all EVTYPE values where there is an injury, fatality, property damage or crop damage.  Any rows with no injury or damage (i.e. a value of 0 or less) will be subsetted out as they aren't need for this evaluation.  This leaves 254633 observations that had a recorded impact in which we are interested.

```{r cleanup EVTYPE data}
weather.data.of.interest <- weather.data[
                                         (
                                          weather.data[, "FATALITIES"] > 0 |
                                          weather.data[, "INJURIES"] > 0 |
                                          weather.data[, "CROPDMG"] > 0 |
                                          weather.data[, "PROPDMG"] > 0
                                         ),
                                        ]
##Astronomical Low Tide
weather.data.of.interest$EVTYPE <- gsub(".*Astronomical Low Tide.*", "Astronomical Low Tide", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Avalanche
weather.data.of.interest$EVTYPE <- gsub(".*mud.*slide.*|.*avalanc[h]*e.*|^rock slide.*|.*landslide.*", "Avalanche", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Cold/WindChill
weather.data.of.interest$EVTYPE <- gsub("^unseasonabl.* cold.*|^cold.*|^extreme.*cold.*|^extreme.*chill.*|fog and cold.*|low temperature", "Cold/Wind Chill", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Dust Devil
weather.data.of.interest$EVTYPE <- gsub("^Dust Devil*", "Dust Devil", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Dust Storm
weather.data.of.interest$EVTYPE <- gsub("^Dust Storm.*|BLowing Dust.*", "Dust Storm", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Excessive Heat
weather.data.of.interest$EVTYPE <- gsub(".*excessive heat.*|.*extreme heat.*", "Excessive Heat", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Extreme Cold/Wind Chill
weather.data.of.interest$EVTYPE <- gsub(".*record.*cold.*", "Extreme Cold/Wind Chill", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Flood
weather.data.of.interest$EVTYPE <- gsub(".*flood.*|.*urban.*|dam break", "Flood", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Fog
weather.data.of.interest$EVTYPE <- gsub("^dense.*|fog", "Fog", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Frost/Freeze
weather.data.of.interest$EVTYPE <- gsub(".*frost.*|.*freezing.*|.*freeze.*", "Frost/Freeze", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Hail
weather.data.of.interest$EVTYPE <- gsub(".*hail.*", "Hail", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Heat
weather.data.of.interest$EVTYPE <- gsub("^unseasonably warm.*|.*drought.*|.*fire.*|^heat.*|^record.*heat.*", "Heat", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Heavy Rain
weather.data.of.interest$EVTYPE <- gsub(".*excessive wetness.*|.*microburst.*|.*torrential rain.*|.*unseasonal rain.*|.*heavy.*rain.*|.*heavy.*mix.*|mixed.*precipitation.*|heavy precipitation|heavy shower|record.*rainfall|.*rain/wind.*|downburst|cool and wet|drowning|rain|mixed precip|coastal.*", "Heavy Rain", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Heavy Snow
weather.data.of.interest$EVTYPE <- gsub("^heavy snow.*.*lake.*effect.*|.*snow.*|.*blizzard.*", "Heavy Snow", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##High Surf
weather.data.of.interest$EVTYPE <- gsub("^rough surf.*|^rough seas.*|^rogue wave.*|.*high.*surf.*|^heavy surf.*|heavy swells|heavy seas|hazardous surf|rapidly rising water", "High Surf", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##High Wind
weather.data.of.interest$EVTYPE <- gsub("^high wind.*|^wind.*|.*high.*wind.*|gustnado|gusty.*wind.*|gradient wind|non-severe wind.*|.*mircoburst.*", "High Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Hurricane (Typhoon)
weather.data.of.interest$EVTYPE <- gsub(".*typhoon.*|.*tsunami.*|.*hurricane.*", "Hurricane (Typhoon)",  weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Ice Storm
weather.data.of.interest$EVTYPE <- gsub(".*ice.*|.*icy.*|glaze", "Ice Storm",  weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Lightning
weather.data.of.interest$EVTYPE <- gsub(".*Lig[h]*[n]*[t]*[n]*ing.*", "Lightning", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Marine Strong Wind
weather.data.of.interest$EVTYPE <- gsub("^Marine mishap.*|marine strong wind|marine accident.*", "Marine Strong Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Marine Thunderstorm Wind
weather.data.of.interest$EVTYPE <- gsub(".*marine thunderstorm wind.*|marine tstm wind", "Marine Thunderstorm Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Rip Current
weather.data.of.interest$EVTYPE <- gsub(".*rip current.*", "Rip Current", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Seiche
weather.data.of.interest$EVTYPE <- gsub("^seiche.*", "Seiche", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Sleet
weather.data.of.interest$EVTYPE <- gsub("^sleet.*", "Sleet", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Strong Wind
weather.data.of.interest$EVTYPE <- gsub("^strong wind.*|^storm force.*|^Severe turbulence.*|non.*tstm.*wind.*", "Strong Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Storm Surge
weather.data.of.interest$EVTYPE <- gsub(".*surge.*|^storm surge.*|high|.*high seas.*|high.*tides.*|.*storm surge.*", "Storm Surge/Tide", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Thunderstorm Wind
weather.data.of.interest$EVTYPE <- gsub(".*tstm.*wind.*|^t[h]*u[n]*[d]*e[e]*r[e]*[s]*torm.*|thunderstorms.* wind[s].*|wins.*|^thunders.*", "Thunderstorm Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
weather.data.of.interest$EVTYPE <- gsub("^severe thunderstorm.*", "Thunderstorm Wind", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Tornado
weather.data.of.interest$EVTYPE <- gsub(".*torndao.*|landspout|^whirlwind.*|.*tornado.*|funnel.*", "Tornado", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Tropical Depression
weather.data.of.interest$EVTYPE <- gsub("^tropical depression.*", "Tropical Depression", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Tropical Storm
weather.data.of.interest$EVTYPE <- gsub("^tstm.*|^tropical storm.*", "Tropical Storm", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Volcanic Ash
weather.data.of.interest$EVTYPE <- gsub("^Volcanic.*", "Volcanic Ash", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Waterspout
weather.data.of.interest$EVTYPE <- gsub(".*waterspout.*", "Waterspout", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Wildfire
weather.data.of.interest$EVTYPE <- gsub("^wildfire.*|.*wild.*fire.*", "Wildfire", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Winter Storm
weather.data.of.interest$EVTYPE <- gsub("^winter |winter winter storm.*|^snow.*|weather.*|.*winter storm.*", "Winter Storm", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Winter Weather
weather.data.of.interest$EVTYPE <- gsub("^hyperthermia.*|hypothermia.*|^light snow.*|^wintry mix.*", "Winter Weather", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Heavy Rain
weather.data.of.interest$EVTYPE <- gsub(".*heavy.*rain.*", "Heavy Rain", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Storm Surge
weather.data.of.interest$EVTYPE <- gsub(".*storm surge.*", "Storm Surge/Tide", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
##Other
weather.data.of.interest$EVTYPE <- gsub("^Other.*|Apache County", "Other", weather.data.of.interest$EVTYPE, ignore.case = TRUE)
```
##Final list of events of interest for this analysis
```{r}
sort(unique(weather.data.of.interest$EVTYPE))
```
##Aggregate the results by EVTYPE
This sesion of code will aggregate the results of the fatalities, injuries and property/crop damage.  This will be used to generate the visuals below to demonstrate the weather events with the most destructive impact.
```{r aggregate results}
##Aggregate by event type
total.fatalities <- aggregate(weather.data.of.interest[, "FATALITIES"], by = list(weather.data.of.interest[, "EVTYPE"]), FUN = sum, na.rm = TRUE)
total.injuries <- aggregate(weather.data.of.interest[, "INJURIES"], by = list(weather.data.of.interest[, "EVTYPE"]), FUN = sum, na.rm = TRUE)

colnames(total.fatalities) <- c("Event Type", "Total")
colnames(total.injuries) <- c("Event Type", "Total")
```
##Assigning values to damage
The data in the PROPDMGEXP and CROPDMGEXP appear to be an exponential.  However, some of the data is short-handed such as H for Hundred (or 10^2), K for Thousand (or 10^3), M for Million (or 10^6) and B for Billion (or 10^9).  The other values we will set to 0 as they seem to be errors.
```{r fix exponents}
##PROPDMGEXP fixes
levels(weather.data.of.interest[, "PROPDMGEXP"]) <- c(unique(levels(weather.data.of.interest[, "PROPDMGEXP"])), 9)
weather.data.of.interest[weather.data.of.interest[, "PROPDMGEXP"] == "", "PROPDMGEXP"] = 0
weather.data.of.interest[
                         (
                           weather.data.of.interest[, "PROPDMGEXP"] == "-"
                          |weather.data.of.interest[, "PROPDMGEXP"] == "+"
                          |weather.data.of.interest[, "PROPDMGEXP"] == "?"
                         ), "PROPDMGEXP"
                        ] = 1
weather.data.of.interest[weather.data.of.interest[, "PROPDMGEXP"] == "H", "PROPDMGEXP"] = 2
weather.data.of.interest[weather.data.of.interest[, "PROPDMGEXP"] == "K", "PROPDMGEXP"] = 3
weather.data.of.interest[weather.data.of.interest[, "PROPDMGEXP"] == "M", "PROPDMGEXP"] = 6
weather.data.of.interest[weather.data.of.interest[, "PROPDMGEXP"] == "B", "PROPDMGEXP"] = 9

##CROPDMGEXP fixes
levels(weather.data.of.interest[, "CROPDMGEXP"]) <- c(unique(levels(weather.data.of.interest[, "CROPDMGEXP"])), 0, 1, 2, 3, 6, 9)
weather.data.of.interest[weather.data.of.interest[, "CROPDMGEXP"] == "", "CROPDMGEXP"] = 0
weather.data.of.interest[
                         (
                           weather.data.of.interest[, "CROPDMGEXP"] == "-"
                          |weather.data.of.interest[, "CROPDMGEXP"] == "+"
                          |weather.data.of.interest[, "CROPDMGEXP"] == "?"
                         ), "CROPDMGEXP"
                        ] = 1
weather.data.of.interest[weather.data.of.interest[, "CROPDMGEXP"] == "H", "CROPDMGEXP"] = 2
weather.data.of.interest[weather.data.of.interest[, "CROPDMGEXP"] == "K", "CROPDMGEXP"] = 3
weather.data.of.interest[weather.data.of.interest[, "CROPDMGEXP"] == "M", "CROPDMGEXP"] = 6
weather.data.of.interest[weather.data.of.interest[, "CROPDMGEXP"] == "B", "CROPDMGEXP"] = 9
```
Aggregate the property damage and crop damage results
```{r aggregate damage results}
##Get the total property and crop damage raised to the imputted exponent
PROPDMGAMOUNT <- weather.data.of.interest[, "PROPDMG"] * (10 ^ as.numeric(weather.data.of.interest[, "PROPDMGEXP"]))
CROPDMGAMOUNT <- weather.data.of.interest[, "CROPDMG"] * (10 ^ as.numeric(weather.data.of.interest[, "CROPDMGEXP"]))
TOTALDMGAMOUNT <- PROPDMGAMOUNT + CROPDMGAMOUNT
##Add the total column to the data frame
weather.data.of.interest <- cbind(weather.data.of.interest, TOTALDMGAMOUNT)
##Compute the aggregate total by event type
total.property.damage <- aggregate(
                                    TOTALDMGAMOUNT
                                   ,by = list(weather.data.of.interest[, "EVTYPE"])
                                   ,FUN = sum
                                   ,na.rm = TRUE
                                  )

colnames(total.property.damage) <- c("Event Type", "Total")
```
##Plot the data to determine worst events to humans and property
```{r plotting the data}
##Get the top 10 values for each plot
top.total.fatalities <- total.fatalities[order(-total.fatalities[, "Total"]), ][1:10, ]
top.total.injuries <- total.injuries[order(-total.injuries[, "Total"]), ][1:10, ]
top.total.property.damage <- total.property.damage[order(-total.property.damage[, "Total"]), ][1:10, ]
```
## Results
The worst event for damage to humans (death and injury) are 1.) Tornados and 2.) Excessive Heat. The worst events for property damage are 1.) Floods and 2.) Hurricanes (Typhoons)

##Pie Chart Generation
```{r generate the charts}
##Generate pie charts
pie(top.total.fatalities[, "Total"], labels = top.total.fatalities[, "Event Type"], main = "Top 10 Fatality Causing Events")
pie(top.total.injuries[, "Total"], labels = top.total.injuries[, "Event Type"], main = "Top 10 Injury Causing Events")
pie(top.total.property.damage[, "Total"], labels = top.total.property.damage[, "Event Type"], main = "Top 10 Property Damage Causing Events")
```


[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2